A Constraint Solving Approach to Parikh Images of Regular Languages

### 2024-03-06 abstract.tex
Fix units, mention SMT-COMP results in abstract

A common problem in string constraint solvers is computing the Parikh image, a
linear arithmetic formula that describes all possible combinations of character
counts in strings of a given language. Automata-based string solvers frequently
need to compute the Parikh image of large products (intersections) of
nondeterministic automata, which in many operations is both prohibitively slow
and memory-intensive. We contribute a novel understanding of how Parikh maps can
be tackled as a constraint solving problem to solve real-world constraints
stemming from functions on regular languages, including length and indexing
constraints. Furthermore, we show how this formulation can be efficiently
implemented as a calculus, \Calculus{}, in an automated theorem prover
supporting Presburger logic.  We implement \Calculus{} in a tool called
\Catra{}, and evaluate it on constraints produced by the \OstrichPlus{} string
constraint solver when solving Parikh automata intersection problems produced
when solving standard string constraint benchmarks involving cardinality
constraints on strings. We show that our solution strictly outperforms the
standard approach described in~\citeauthor{generate-parikh-image} as well as the
over-approximating method recently described by~\citeauthor{approximate-parikh}
by a wide margin, and for realistic timeouts for constraint solving also
the~\Nuxmv{} model checker.

>>>>

A common problem in string constraint solvers is computing the Parikh image, a
linear arithmetic formula that describes all possible combinations of character
counts in strings of a given language. Automata-based string solvers frequently
need to compute the Parikh image of large products (intersections) of
nondeterministic automata, which in many operations is both prohibitively slow
and memory-intensive. We contribute a novel understanding of how Parikh maps can
be tackled as a constraint solving problem to solve real-world constraints
stemming from functions on regular languages, including length and indexing
constraints. Furthermore, we show how this formulation can be efficiently
implemented as a calculus, \Calculus{}, in an automated theorem prover
supporting Presburger logic.  We implement \Calculus{} in a tool called
\Catra{}, and evaluate it on constraints produced by the \OstrichPlus{} string
constraint solver when solving Parikh automata intersection problems produced
when solving standard string constraint benchmarks involving cardinality
constraints on strings. We show that our solution strictly outperforms the
standard approach described by~\citeauthor{generate-parikh-image} as well as the
over-approximating method recently described by~\citeauthor{approximate-parikh}
by a wide margin, and for realistic timeouts (under \SI{60}{s}) also
the~\Nuxmv{} model checker. When added as the Parikh image backend of \OstrichPlus{} to
the \Ostrich{} string constraint solver's portfolio, it boosts its results on the
quantifier-free strings with linear integer algebra track of SMT-COMP~20203 (QF\_SLIA)
enough to solve the most \Unsat{} instances in that track of all competitors.

### 2024-03-06 evaluation.tex
Tables: caption above
Fix units, mention SMT-COMP results in abstract

The benchmarks are executed in parallel across a cluster of identical
machines, \BenchmarkRig{}, as the sole job for each machine.  We compiled
the code using Scala~\ScalaVersion{} and executed the experiments
on~\JvmVersion{} with a maximum heap of~\MaxHeapSize{}. We used \Nuxmv{}
version~\NuxmvVersion{} invoked as a subprocess for each instance. Instances
were executed in batches of \BatchSize{}, each given a fresh JVM. Each JVM was
warmed up for \numprint{10}s on a random benchmark from the set before starting
to execute. We believe this represents a realistic use case where \Calculus{} is
used to support, e.g., a string solver. Experiments were executed in random order
for all backends. Each instance got a time budget of~\RuntimeTimeout{}.

All runtimes are measured in wall-clock time as observed by the JVM when
executing the instance, and exclude time spent parsing (usually far below
\numprint{0.1}s).

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-by-solver.pdf}
    \caption{The division of statuses per backend.}
    \label{fig:solve-division}
    \Description[A bar chart showing three bars, one per backend, illustrating how many instances they could solve]{The bars are divided by satisfiable and unsatisfiable instances. Baseline could seemingly only solve unsatisfiable instances, PC* could solve a few satisfiable and mostly unsatisfiable, and nuxmv could solve about twice as many unsatisfiable as satisfiable instances, and slightly less in total than PC*.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-time-boxplot.pdf}
    \caption{The distribution of runtimes for solved instances per backend, assuming they were solved within the \numprint{30}s timeout.
    The whiskers use the default placement in Pandas, \numprint{1.5} times the IQR (interquartile range).}
    \label{fig:runtime-boxplot}
    \Description[A box plot showing the distribution of runtime over the three backends]{The middle box plot shows a tiny box centered around 0 seconds for the PC* backend, the rightmost box shows a bigger box between five seconds and 30 seconds for nuxmv, and the leftmost box shows a smaller box between 20 and 30 seconds for baseline. The whiskers of nuxmv span between 0 and the timeout, 60 seconds, while they are much tighter for PC*. On the other hand, PC* has a lot of outliers.}
  \end{subfigure}
  \vfill
  \begin{subfigure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-cactus.pdf}
    \caption{The number of instances solved as the time budget increases, simulated from one \numprint{120}s-timeout run.}
    \label{fig:cactus}
    \Description[A cactus plot comparing the performance of nuxmv, PC* and baseline]{The line for baseline is strictly lower than the two others, and does not head upwards from zero until after 10 seconds of timeout, while the number of instances for nuxmv increases sharply until ca 10 seconds and then linearly after that. Lazy solves much more instances than nuxmv.}
  \end{subfigure}
  \caption{Evaluation of \Calculus{} solving Parikh automata problems generated by \Ostrich{}.}
\end{figure}

@@

Finally, in \cref{fig:runtime-boxplot,fig:duels:baseline,fig:duels:nuxmv} and we compare
runtimes for solved instances between the backends. We see that \Nuxmv{} has a more even spread of runtimes up to \numprint{20}s,
while both \Calculus{} and baseline tend towards solving their
instances quickly or not at all, though \Calculus{} does have a long tail of outlier
instances that finish as the timeout increaseses.
Notably, \cref{fig:duels:baseline} shows that any instance solved by baseline in \numprint{30}s is also
solved by \Calculus{} in under \numprint{5}s.

\begin{table}
  \begin{center}
  \input{graphs/solved_pivot_table.tex}
  \end{center}
  \caption{Number of successful results within a timeout of \RuntimeTimeout{}.
  Instances solved by no backend within the timeout (about half of the set) are
  omitted from the table.}\label{tab:solve-status}
\end{table}

@@

\begin{table}
  \begin{center}
  \input{graphs/qf_slia_comparison.tex}
  \end{center}
  \caption{Number of solved benchmarks in the set of quantifier-free strings with
    linear integer arithmetic constraints (QF\_SLIA) at SMT-COMP~2023. The numbers are from
    the competition results, except for CA-Str, which is executed by us on the same 
    cluster as the competition with the same resources, and the two virtual portfolio solvers
    \Ostrich+CA and \textsc{Competition} which aggregate the best results from \Ostrich{}/CA-Str and 
    all the competition results for non-\Ostrich{} solvers respectively. }
  \label{tab:solve-status-smt-comp}
\end{table}

>>>>

The benchmarks are executed in parallel across a cluster of identical
machines, \BenchmarkRig{}, as the sole job for each machine.  We compiled
the code using Scala~\ScalaVersion{} and executed the experiments
on~\JvmVersion{} with a maximum heap of~\MaxHeapSize{}. We used \Nuxmv{}
version~\NuxmvVersion{} invoked as a subprocess for each instance. Instances
were executed in batches of \BatchSize{}, each given a fresh JVM. Each JVM was
warmed up for \SI{10}{s} on a random benchmark from the set before starting
to execute. We believe this represents a realistic use case where \Calculus{} is
used to support, e.g., a string solver. Experiments were executed in random order
for all backends. Each instance got a time budget of~\RuntimeTimeout{}.

All runtimes are measured in wall-clock time as observed by the JVM when
executing the instance, and exclude time spent parsing (usually far below
\SI{0.1}{s}).

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-by-solver.pdf}
    \caption{The division of statuses per backend.}
    \label{fig:solve-division}
    \Description[A bar chart showing three bars, one per backend, illustrating how many instances they could solve]{The bars are divided by satisfiable and unsatisfiable instances. Baseline could seemingly only solve unsatisfiable instances, PC* could solve a few satisfiable and mostly unsatisfiable, and nuxmv could solve about twice as many unsatisfiable as satisfiable instances, and slightly less in total than PC*.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-time-boxplot.pdf}
    \caption{The distribution of runtimes for solved instances per backend, assuming they were solved within the \SI{30}{s} timeout.
    The whiskers use the default placement in Pandas, \numprint{1.5}~times the IQR (interquartile range).}
    \label{fig:runtime-boxplot}
    \Description[A box plot showing the distribution of runtime over the three backends]{The middle box plot shows a tiny box centered around 0 seconds for the PC* backend, the rightmost box shows a bigger box between five seconds and 30 seconds for nuxmv, and the leftmost box shows a smaller box between 20 and 30 seconds for baseline. The whiskers of nuxmv span between 0 and the timeout, 60 seconds, while they are much tighter for PC*. On the other hand, PC* has a lot of outliers.}
  \end{subfigure}
  \vfill
  \begin{subfigure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-cactus.pdf}
    \caption{The number of instances solved as the time budget increases, simulated from one \SI{120}{s}-timeout run.}
    \label{fig:cactus}
    \Description[A cactus plot comparing the performance of nuxmv, PC* and baseline]{The line for baseline is strictly lower than the two others, and does not head upwards from zero until after 10 seconds of timeout, while the number of instances for nuxmv increases sharply until ca 10 seconds and then linearly after that. Lazy solves much more instances than nuxmv.}
  \end{subfigure}
  \caption{Evaluation of \Calculus{} solving Parikh automata problems generated by \Ostrich{}.}
\end{figure}

@@

Finally, in \cref{fig:runtime-boxplot,fig:duels:baseline,fig:duels:nuxmv} and we compare
runtimes for solved instances between the backends. We see that \Nuxmv{} has a more even spread of runtimes up to \SI{20}{s},
while both \Calculus{} and baseline tend towards solving their
instances quickly or not at all, though \Calculus{} does have a long tail of outlier
instances that finish as the timeout increaseses.
Notably, \cref{fig:duels:baseline} shows that any instance solved by baseline in \SI{30}{s} is also
solved by \Calculus{} in under \SI{5}{s}.

\begin{table}
  \caption{Number of successful results within a timeout of \RuntimeTimeout{}.
  Instances solved by no backend within the timeout (about half of the set) are
  omitted from the table.}\label{tab:solve-status}
  \begin{center}
  \input{graphs/solved_pivot_table.tex}
  \end{center}
\end{table}

@@

\begin{table}
  \caption{Number of solved benchmarks in the set of quantifier-free strings with
  linear integer arithmetic constraints (QF\_SLIA) at SMT-COMP~2023. The numbers are from
  the competition results, except for CA-Str, which is executed by us on the same 
  cluster as the competition with the same resources, and the two virtual portfolio solvers
  \Ostrich+CA and \textsc{Competition} which aggregate the best results from \Ostrich{}/CA-Str and 
  all the competition results for non-\Ostrich{} solvers respectively. }\label{tab:solve-status-smt-comp}
  \begin{center}
  \input{graphs/qf_slia_comparison.tex}
  \end{center}
\end{table}

### 2024-03-06 extensions.tex
Clarify backward propagation, dom()
Closes: #81
Fix units, mention SMT-COMP results in abstract

\Calculus{} can be accelerated for some instances by adding rules for
backjumping. In particular, key connectedness constraints for an automaton can
be learnt. In that case, when discovering that a transition $\Transition$ of an
automaton $\Automaton$ under a certain transition variable $\Filter$ has become
unreachable, we can learn the clause $\Connected(\Automaton, \Filter) \land
\bigwedge_{\Transition' \in C} \Filter(\Transition') = 0 \implies \Transition = 0$ where
$\Dominates(C, \Automaton, \Transition)$ for some minimal $C$. Note that this
captures both forward (unrachable from the initial state) and
backward (does not reach an accepting state) unreachability. Bidirectional cut
learning is implemented in \Catra, and has provided a slight improvement in
performance on some instances.

@@

\iffalse
\subsection{Transducers}\label{sec:ext:transducers}

Another interesting application is applying \Calculus{} to automata with
multiple tracks, e.g. transducers. One application of such a calculus would be
to represent replace operations and other functions on regular languages, and to
be able to answer questions like "does this operation change the length of the
string". The difficulty in implementing it for transducers is first to perform
the mapping on the labels, which for both the length and Parikh cases is
straightforward; just do the same thing in two dimensions. The more complicated
operation is defining the product of transducers. In some cases it would
probably be desirable to perform synchronisation (e.g. requiring overlapping
transitions) on only some transitions, for example the first track.

@@

    A preliminary implementation of this approach is available as part
    of~\Catra, but has been neither optimised nor fully tested.
\fi

>>>>

\Calculus{} can be accelerated for some instances by adding rules for
backjumping. In particular, key connectedness constraints for an automaton can
be learnt. In that case, when discovering that a transition $\Transition$ of an
automaton $\Automaton$ under a certain transition variable $\Filter$ has become
unreachable, we can learn the clause $\Connected(\Automaton, \Filter) \land
\bigwedge_{\Transition' \in C} \Filter(\Transition') = 0 \implies \Transition = 0$ where
$\Dominates(C, \Automaton, \Transition)$ for some minimal $C$. Note that this
captures both forward (unreachable from the initial state) and
backward (does not reach an accepting state) unreachability. Bidirectional cut
learning is implemented in \Catra, and has provided a slight improvement in
performance on some instances.

@@

\subsection{Transducers}\label{sec:ext:transducers}

Another interesting application is applying \Calculus{} to automata with
multiple tracks, e.g.\ transducers. One application of such a calculus would be
to represent replace operations and other functions on regular languages, and to
be able to answer questions such as \enquote{does this operation change the length of the
string}. The difficulty in implementing it for transducers is first to perform
the mapping on the labels, which for both the length and Parikh cases is
straightforward; just do the same thing in two dimensions. The more complicated
operation is defining the product of transducers. In some cases it would
probably be desirable to perform synchronisation (e.g.\ requiring overlapping
transitions) on only some transitions, for example the first track.

@@

    A preliminary implementation of this approach is available as part
    of~\Catra, but has been neither optimised nor fully tested.

### 2024-03-06 introduction.tex
Fix units, mention SMT-COMP results in abstract

We additionally wrap the Parikh image solver as a stand-alone tool, \Catra{}.
\Catra{} also supports a variant of the approximate method of~\cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra, we compare \Calculus{} to the other two
back-ends on \NrBenchmarks{} distinct Parikh automata intersection problems
generated by \OstrichPlus{} when solving the PyEx string constraint benchmark
suite involving string length constraints~\cite{pyex}, finding that \Calculus{}
outperforms both \Nuxmv{} and the baseline method, allowing \Calculus{}
to solve every problem solved by baseline within \numprint{30}s in under \numprint{5}s.
\Calculus{} in particular outperforms \Nuxmv{} on unsatisfiable instances, more than 
doubling the number of unsatisfiable results found within a \numprint{30}s timeout.

>>>>

We additionally wrap the Parikh image solver as a stand-alone tool, \Catra{}.
\Catra{} also supports a variant of the approximate method of~\cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra, we compare \Calculus{} to the other two
back-ends on \NrBenchmarks{} distinct Parikh automata intersection problems
generated by \OstrichPlus{} when solving the PyEx string constraint benchmark
suite involving string length constraints~\cite{pyex}, finding that \Calculus{}
outperforms both \Nuxmv{} and the baseline method, allowing \Calculus{}
to solve every problem solved by baseline within \SI{30}{s} in under \SI{5}{s}.
\Calculus{} in particular outperforms \Nuxmv{} on unsatisfiable instances, more than 
doubling the number of unsatisfiable results found within a \SI{30}{s}~timeout.
